---
title: "10_AGF_juvenile_Akenti_survival_statistics_ bayes"
author: "Alex Macadam"
date: "2023-10-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
set.seed(123)
library(tidyverse) #for data wrangling
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(brms)
library(broom.mixed)
library(tidybayes)
library(bayesplot)
library(rstanarm)
library(cmdstanr)
library(ggeffects)
library(rstan)
library(DHARMa)
library(ggridges)
library(easystats)
library(patchwork)
source('/Users/alexmacadam/Documents/GitHub/AGF-Physiology/helperFunctions.R')
```

Survival
```{r}
#load data
survival <- read.csv("/Users/alexmacadam/Documents/GitHub/AGF-Physiology/data/Juvenile-survival-2020.csv", header=T, fileEncoding="UTF-8-BOM") |>
  mutate(Family = factor(Family),
         Species = factor(Species),
         Temp = factor(Temp),
         Tank = factor(Tank),
         Timepoint = factor(Timepoint),
         Cassette = factor(Cassette),
         Plate.Slide = factor(Plate.Slide),
         Treatment = factor(Treatment),
         Cross = factor(Cross),
         Crosstype = factor(Crosstype),
         RegionCross = factor(RegionCross),
         popcross = factor(popcross)
         ) |>
  filter(Timepoint == "6", Crosstype == "Individual Cross")

survival.hot <- survival |> filter(Temp == "HOT")
```

```{r}
fams <- c("Jewell8xDavies13", "Jewell8xPalms5", "Jewell8xJewell2", "Jewell8xJewell3", "Jewell8xDavies1", "Jewell8xDavies6")
survival.hot.fams <- survival.hot |> dplyr::filter(Cross %in% fams)
```

```{r}
# PERCENT SURVIVAL
# Function for summarising survival standard error
summarySE <- function(data=NULL, measurevar=NULL, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
  require(plyr)
  
  # New version of length which can handle NA's: if na.rm==T, don't count them
  length2 <- function (x, na.rm=FALSE) {
    if (na.rm==TRUE) sum(!is.na(x))
    else       length(x)
  }
  
  # This does the summary. For each group's data frame, return a vector with
  # N, mean, and sd
  datac <- ddply(data, groupvars, .drop=.drop,
                 .fun = function(xx, col) {
                   c(N    = length2(xx[[col]], na.rm=na.rm),
                     mean = mean   (xx[[col]], na.rm=na.rm),
                     sd   = sd     (xx[[col]], na.rm=na.rm)
                   )
                 },
                 measurevar
  )
  
  # Commenting this rename mean column line, as not assigning measurevar name properly
  # # Rename the "mean" column    
  # datac <- rename(datac, measurevar = mean)
  
  datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean
  
  # Confidence interval multiplier for standard error
  # Calculate t-statistic for confidence interval: 
  # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
  ciMult <- qt(conf.interval/2 + .5, datac$N-1)
  datac$ci <- datac$se * ciMult
  
  return(datac)
}
```

```{r}
surv.summ <- summarySE(survival, measurevar="Survival",
                     groupvars=c("Temp"), na.rm=TRUE)
```

# Fit the model
form
```{r}
survival.form <- bf(Survival ~ Treatment + (1|Tank),
                  family=bernoulli(link='logit'))
```

priors
```{r fitModel2h1, results='markdown', eval=TRUE, mhidden=TRUE, cache=TRUE}
priors <-
    prior(normal(1, 1.7), class = 'Intercept') +
    prior(normal(0, 1.7), class = 'b') +
    prior(student_t(3, 0, 1.7), class = 'sd') 
```

run model
```{r}
survival.brm <- brm(survival.form, 
                  data = survival.hot.fams,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 5000,
                  warmup = 2500,
                  chains = 3, cores = 3,
                  thin = 5,
                  control = list(adapt_delta = 0.99, max_treedepth = 20),
                  refresh = 0
                  )
```

View prior and posterior
```{r}
survival.brm %>% SUYR_prior_and_posterior()
```

Trace plots
```{r}
survival.brm$fit |> stan_trace()
```

Autocorrelation
```{r}
survival.brm$fit |> stan_ac()
```

Rhat
```{r}
survival.brm$fit |> stan_rhat()
```

Effective sample size
```{r}
survival.brm$fit |> stan_ess()
```

Denisty overlay
```{r}
survival.brm |> pp_check(type = 'dens_overlay', nsamples = 250)
```

Dharma residuals
```{r}
survival.resids<- make_brms_dharma_res(survival.brm, integerResponse = FALSE)
wrap_elements(~testUniformity(survival.resids)) +
wrap_elements(~plotResiduals(survival.resids, form = factor(rep(1,nrow(survival.hot))))) +
wrap_elements(~plotResiduals(survival.resids, quantreg = FALSE))+
wrap_elements(~testDispersion(survival.resids))

survival.resids<- make_brms_dharma_res(survival.brm, integerResponse = FALSE)
testUniformity(survival.resids)
plotResiduals(survival.resids, form = factor(rep(1,nrow(survival.hot))))
plotResiduals(survival.resids, quantreg = FALSE)
testDispersion(survival.resids)
```

Summarise model
conditional effects
```{r}
survival.brm %>%
    conditional_effects() %>%
    plot(points = TRUE)
```

Planned contrast:
D1 hot
Sed hot
SS hot
```{r}
cmat=cbind(
  "D1vSS" = c(1, 0, -1),
  "D1vSed" = c(1, -1, 0),
  "SSvSed" = c(0, -1, 1)
)
```

```{r}
survival.em<- survival.brm |>
  emmeans(~Treatment, type = 'response') |>
  contrast(method=list(cmat)) |>
  gather_emmeans_draws() |>
  mutate(across(everything(), exp)) |>
#on odds ratio scale
  dplyr::summarise(median_hdci(.value),
            Pl = mean(.value < 1),
            Pg = mean(.value > 1)
            )
survival.em
```

#Plot
```{r}
surv.plot<- survival.brm %>%
  emmeans(~Treatment) %>%
  gather_emmeans_draws() %>%
    ggplot() +
  #geom_vline(xintercept = 0, linetype = "dashed") +
  stat_slab(aes(
    x = .value, y = Treatment,
    fill = stat(ggdist::cut_cdf_qi(cdf,
      .width = c(0.5, 0.8, 0.95),
      labels = scales::percent_format()
    ))
  ), color = "black") +
  scale_fill_brewer("Interval", direction = -1, na.translate = FALSE) +
  theme_classic() +
  labs(x = "Survival", y = "Treatment")
surv.plot
```

######
Size
```{r}
#load data
size <- read.csv("/Users/alexmacadam/Documents/GitHub/AGF-Physiology/data/Juvenile-size-2020.csv", header=T, fileEncoding="UTF-8-BOM") |>
  mutate(Family = factor(Family),
         Species = factor(Species),
         Temp = factor(Temp),
         Tank = factor(Tank),
         Timepoint = factor(Timepoint),
         Timepoint2 = factor(Timepoint2),
         Cassette = factor(Cassette),
         Plate.Slide = factor(Plate.Slide),
         Treatment = factor(Treatment, levels = c("D1", "SS", "Jewell")),
         Cone = factor(Cone),
         Cross = factor(Cross),
         CrossType = factor(CrossType),
         popcross = factor(popcross),
         SINGLE.CLUMP = factor(SINGLE.CLUMP),
         Mom = factor(Mom),
         Dad = factor(Dad)
         ) |>
  filter(CrossType == "IndivCross", SINGLE.CLUMP == "S")

#create individual ID
size$IndID <- paste(size$Tank, size$Treatment, size$Cassette, size$Plate.Slide, size$MatchID,
                        sep='_')

#Add Temp x Treatment column
size$popcrossxsym <- paste(size$Treatment, size$popcross, sep='_')
size$popcrossxsym <- as.factor(size$popcrossxsym)
```

```{r}
fams <- c("Jewell8xDavies13", "Jewell8xPalms5", "Jewell8xJewell2", "Jewell8xJewell3", "Jewell8xDavies1", "Jewell8xDavies6")
size.fams <- size |> dplyr::filter(Cross %in% fams)
```

Growth data
```{r}
growth<- size.fams %>% filter(Timepoint2 %in% c("T0","T6")) %>% dplyr::select(Treatment, Timepoint2, Temp,  Cross, Area, IndID, popcross, Tank, Cassette, Plate.Slide, Mom, Dad) %>%
  mutate(popcross = factor(popcross)) %>%
  pivot_wider(names_from = Timepoint2, values_from = Area) %>%
  tidyr::unnest() %>%
  mutate(Growth = ((`T6` - `T0`)/`T0`)*100) |>
  drop_na(Growth)

growth$popcrossxsym <- paste(growth$Treatment, growth$popcross, sep='_')
growth$popcrossxsym <- as.factor(growth$popcrossxsym)
growth$symxtemp <- paste(growth$Treatment, growth$Temp, sep='_')
growth$symxtemp <- as.factor(growth$symxtemp)
```

# Fit the model
form
```{r}
growth.form <- bf(Growth ~ symxtemp + (1|Plate.Slide),
                     family = gaussian()
                   )
```

priors
```{r fitModel2h1, results='markdown', eval=TRUE, hidden=TRUE, cache=TRUE}
priors<- prior("normal(-10, 5)", class = "Intercept") +
    prior("normal(0, 15)", class = "b") +
    #prior("normal(0, 0.5)", class = "sd", group = "Tank") +
    prior("normal(0, 7)", class = "sd", group = "Plate.Slide") +
    prior(student_t(2, -5, 1), class = 'sigma') +
    prior(cauchy(0,2.5), class = 'sd')
```

run model
```{r}
growth.brm2 <- brm(growth.form, 
                  data = growth,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 5000,
                  warmup = 1000,
                  chains = 3,
                  thin = 5,
                  control = list(adapt_delta = 0.99),
                  refresh = 0
                  )
```

View prior and posterior
```{r}
growth.brm2 %>% SUYR_prior_and_posterior()
```

Trace plots
```{r}
growth.brm2$fit |> stan_trace()
```

Autocorrelation
```{r}
growth.brm2$fit |> stan_ac()
```

Rhat
```{r}
growth.brm2$fit |> stan_rhat()
```

Effectiev sample size
```{r}
growth.brm2$fit |> stan_ess()
```

Density overlay
```{r modelValidation5b, results='markdown', eval=TRUE, hidden=TRUE, fig.width=6, fig.height=4}
growth.brm2 %>% pp_check(type = 'dens_overlay', nsamples = 250)
```

DHARMa residuals
```{r}
growth.resids<- make_brms_dharma_res(growth.brm2, integerResponse = FALSE)
wrap_elements(~testUniformity(growth.resids)) +
wrap_elements(~plotResiduals(growth.resids, form = factor(rep(1,nrow(growth))))) +
wrap_elements(~plotResiduals(growth.resids, quantreg = FALSE))+
wrap_elements(~testDispersion(growth.resids))
```

Summarise model
conditional effects
```{r}
growth.brm2 |>
  emmeans(~symxtemp) |>
  as.data.frame() |>
  ggplot(aes(y = emmean, x = symxtemp)) +
  geom_pointrange(aes(ymin = lower.HPD, ymax = upper.HPD))
```

planned contrasts- compare situations
                 D1 vs Sed       D1 vs SS        SS vs Sed       Central vs Hybrid       
D1_AMB     |      0                -1/2                0                   -1/3
D1_HOT     |     -1                -1/2                0                    1/4
Jewell_AMB |      0              0                  1/2                 -1/3
Jewell_HOT |      0              0                  1/2                  1/4
SS_AMB     |      0                1/2               -1/2                 -1/3
SS_HOT     |      1                1/2               -1/2                  1/4
 
```{r}
cmat<- cbind(
             "D1_HOT vs Sed_HOT"=c(0, -1, 0, 1, 0, 0),
             "D1_HOT vs SS_HOT" = c(0, -1, 0, 0, 0, 1),
             "SS_HOT vs Sed_HOT" = c(0, 0, 0, 1, 0, -1),
             "D1_AMB vs Sed_AMB"=c(-1, 0, 1, 0, 0, 0),
             "D1_AMB vs SS_AMB" = c(-1, 0, 0, 0, 1, 0),
             "SS_AMB vs Sed_AMB" = c(0, 0, 1, 0, -1, 0),
             "D1_AMB vs D1_HOT" = c(1, -1, 0, 0, 0, 0),
             "SS_AMB vs SS_HOT" = c(0, 0, 0, 0, 1, -1),
             "SED_AMB vs SED_HOT" = c(0, 0, 1, -1, 0, 0),
             "Amb vs hot" = c(-1/3, 1/3, -1/3, 1/3, -1/3, 1/3)
             )
```
 
```{r}
growth.brm2 |>
  emmeans(~symxtemp) |>
  contrast(list(symxtemp=cmat)) |>
  gather_emmeans_draws() |>
  dplyr::summarise(median_hdci(.value),
            Pl = mean(.value < 0),
            Pg = mean(.value > 0)
            )
```

```{r}
growth.plot<- growth.brm2 %>%
  emmeans(~symxtemp) %>%
  gather_emmeans_draws() %>%
    ggplot() +
  #geom_vline(xintercept = 0, linetype = "dashed") +
  stat_slab(aes(
    x = .value, y = symxtemp,
    fill = stat(ggdist::cut_cdf_qi(cdf,
      .width = c(0.5, 0.8, 0.95),
      labels = scales::percent_format()
    ))
  ), color = "black") +
  scale_fill_brewer("Interval", direction = -1, na.translate = FALSE) +
  theme_classic() +
  labs(x = "Growth", y = "Treatment")
growth.plot
```

######
final size
```{r}
final.size<- size |> filter(Timepoint == "6")
final.size$symxtemp <- paste(final.size$Treatment, final.size$Temp, sep='_')
final.size$symxtemp <- as.factor(final.size$symxtemp)
```

form
```{r}
size.form <- bf(Area ~ symxtemp + (1|Plate.Slide),
                     family = gaussian()
                   )
```

priors
```{r fitModel2h1, results='markdown', eval=TRUE, hidden=TRUE, cache=TRUE}
priors<- prior("normal(1, 0.5)", class = "Intercept") +
    prior("normal(0, 0.25)", class = "b") +
    #prior("normal(0, 0.5)", class = "sd", group = "Tank") +
    prior("normal(-1, 1)", class = "sd", group = "Plate.Slide") +
    prior(student_t(3, 0, 1), class = 'sigma') +
    prior(cauchy(0,2.5), class = 'sd')
```

run model
```{r}
size.brm2 <- brm(size.form, 
                  data = final.size,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 2000,
                  warmup = 500,
                  chains = 3,
                  thin = 5,
                  control = list(adapt_delta = 0.99),
                  refresh = 50
                  )
```

View prior and posterior
```{r}
size.brm2 %>% SUYR_prior_and_posterior()
```

Trace plots
```{r}
size.brm2$fit |> stan_trace()
```

Autocorrelation
```{r}
size.brm2$fit |> stan_ac()
```

Rhat
```{r}
size.brm2$fit |> stan_rhat()
```

Effective sample size
```{r}
size.brm2$fit |> stan_ess()
```

Denisty overlay
```{r}
size.brm2 |> pp_check(type = 'dens_overlay', nsamples = 250)
```

Dharma residuals
```{r}
survival.resids<- make_brms_dharma_res(size.brm2, integerResponse = FALSE)
wrap_elements(~testUniformity(survival.resids)) +
wrap_elements(~plotResiduals(survival.resids, form = factor(rep(1,nrow(survival.hot))))) +
wrap_elements(~plotResiduals(survival.resids, quantreg = FALSE))+
wrap_elements(~testDispersion(survival.resids))

survival.resids<- make_brms_dharma_res(size.brm2, integerResponse = FALSE)
testUniformity(survival.resids)
plotResiduals(survival.resids, form = factor(rep(1,nrow(survival.hot))))
plotResiduals(survival.resids, quantreg = FALSE)
testDispersion(survival.resids)
```

Summarise model
conditional effects
```{r partialPlot2d, results='markdown', eval=TRUE, hidden=TRUE, fig.width=8, fig.height=5}
size.brm2 %>%
    conditional_effects() %>%
    plot(points = TRUE)
```

```{r}
size.brm2 |>
  emmeans(~ symxtemp) |>
  gather_emmeans_draws() |>
  dplyr::group_by(symxtemp) |>
  dplyr::summarise(
    median = median(.value),
    lower = quantile(.value, 0.025),
    upper = quantile(.value, 0.975)
  )
```

planned contrasts- compare situations
                 D1 vs Sed       D1 vs SS        SS vs Sed       Central vs Hybrid       
D1_AMB     |      0                -1/2                0                   -1/3
D1_HOT     |     -1                -1/2                0                    1/4
Jewell_AMB |      0              0                  1/2                 -1/3
Jewell_HOT |      0              0                  1/2                  1/4
SS_AMB     |      0                1/2               -1/2                 -1/3
SS_HOT     |      1                1/2               -1/2                  1/4
 
```{r}
cmat<- cbind(
             "D1_HOT vs Sed_HOT"=c(0, -1, 0, 1, 0, 0),
             "D1_HOT vs SS_HOT" = c(0, -1, 0, 0, 0, 1),
             "SS_HOT vs Sed_HOT" = c(0, 0, 0, 1, 0, -1),
             "D1_AMB vs Sed_AMB"=c(-1, 0, 1, 0, 0, 0),
             "D1_AMB vs SS_AMB" = c(-1, 0, 0, 0, 1, 0),
             "SS_AMB vs Sed_AMB" = c(0, 0, 1, 0, -1, 0),
             "D1_AMB vs D1_HOT" = c(1, -1, 0, 0, 0, 0),
             "SS_AMB vs SS_HOT" = c(0, 0, 0, 0, 1, -1),
             "SED_AMB vs SED_HOT" = c(0, 0, 1, -1, 0, 0)
             )
```
 
```{r}
size.brm2 |>
  emmeans(~symxtemp) |>
  contrast(list(symxtemp=cmat)) |>
  gather_emmeans_draws() |>
  dplyr::summarise(median_hdci(.value),
            Pl = mean(.value < 0),
            Pg = mean(.value > 0)
            )
```

```{r}
size.plot<- size.brm2 %>%
  emmeans(~symxtemp) %>%
  gather_emmeans_draws() %>%
    ggplot() +
  #geom_vline(xintercept = 0, linetype = "dashed") +
  stat_slab(aes(
    x = .value, y = symxtemp,
    fill = stat(ggdist::cut_cdf_qi(cdf,
      .width = c(0.5, 0.8, 0.95),
      labels = scales::percent_format()
    ))
  ), color = "black") +
  scale_fill_brewer("Interval", direction = -1, na.translate = FALSE) +
  theme_classic() +
  labs(x = "Size", y = "Treatment")
size.plot
```

##############
Colour
```{r readData, results='markdown', eval=TRUE}
colour <- read.csv("/Users/alexmacadam/Documents/GitHub/AGF-Physiology/data/Juvenile-colour-2020.csv", header=T, fileEncoding="UTF-8-BOM") |>
  mutate( Treatment = factor(Treatment),
          Plate.Slide = factor(Plate.Slide),
          Temp = factor(Temp),
          popcross = factor(popcross)
          ) |>
  filter(IndID != "NA")

colour$symxtemp <- paste(colour$Treatment, colour$Temp, sep='_')
colour$symxtemp <- as.factor(colour$symxtemp)

#colour change
colour<- colour |>
  mutate(T0_2 = (100-(T0 / 255)*100))
colour<- colour |>
  mutate(T6_2 = (100-(T6 / 255)*100))
colour<- colour |>
  mutate(Grey_change_2 = T6_2 - T0_2)
```

# Fit the model
form
```{r}
colour.form <- bf(Grey_change_2 ~ symxtemp + (1|Plate.Slide),
                     family = gaussian()
                   )
```

priors
```{r fitModel2h1, results='markdown', eval=TRUE, hidden=TRUE, cache=TRUE}
priors<- prior("normal(-4, 6)", class = "Intercept") +
    prior("normal(0, 8)", class = "b") +
    #prior("normal(0, 0.5)", class = "sd", group = "Tank") +
    prior("normal(0, 7)", class = "sd", group = "Plate.Slide") +
    prior(student_t(3, 0, 6), class = 'sigma') +
    prior(cauchy(0,2), class = 'sd')
```

run model
```{r}
colour.brm2 <- brm(colour.form, 
                  data = colour,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 5000,
                  warmup = 1000,
                  chains = 3,
                  thin = 10,
                  control = list(adapt_delta = 0.9, max_treedepth = 20),
                  refresh = 0
                  )
```

View prior and posterior
```{r}
colour.brm2 %>% SUYR_prior_and_posterior()
```

Trace plots
```{r}
colour.brm2$fit |> stan_trace()
```

Autocorrelation
```{r}
colour.brm2$fit |> stan_ac()
```

Rhat
```{r}
colour.brm2$fit |> stan_rhat()
```

Effective sample size
```{r}
colour.brm2$fit |> stan_ess()
```

Density overlay
```{r modelValidation5b, results='markdown', eval=TRUE, hidden=TRUE, fig.width=6, fig.height=4}
colour.brm2 %>% pp_check(type = 'dens_overlay', nsamples = 250)
```

DHARMa residuals
```{r}
colour.resids<- make_brms_dharma_res(colour.brm2, integerResponse = FALSE)
wrap_elements(~testUniformity(colour.resids)) +
wrap_elements(~plotResiduals(colour.resids, form = factor(rep(1,nrow(colour))))) +
wrap_elements(~plotResiduals(colour.resids, quantreg = FALSE))+
wrap_elements(~testDispersion(colour.resids))
```

Summarise model
conditional effects
```{r partialPlot2d, results='markdown', eval=TRUE, hidden=TRUE, fig.width=8, fig.height=5}
colour.brm2 %>%
    conditional_effects("symxtemp") %>%
    plot(points = TRUE)
```

planned contrasts- compare situations
                 D1 vs Sed       D1 vs SS        SS vs Sed       Central vs Hybrid       
D1_AMB     |      0                -1/2                0                   -1/3
D1_HOT     |     -1                -1/2                0                    1/4
Jewell_AMB |      0              0                  1/2                 -1/3
Jewell_HOT |      0              0                  1/2                  1/4
SS_AMB     |      0                1/2               -1/2                 -1/3
SS_HOT     |      1                1/2               -1/2                  1/4
 
```{r}
cmat<- cbind(
             "D1_HOT vs Sed_HOT"=c(0, -1, 0, 1, 0, 0),
             "D1_HOT vs SS_HOT" = c(0, -1, 0, 0, 0, 1),
             "SS_HOT vs Sed_HOT" = c(0, 0, 0, 1, 0, -1),
             "D1_AMB vs Sed_AMB"=c(-1, 0, 1, 0, 0, 0),
             "D1_AMB vs SS_AMB" = c(-1, 0, 0, 0, 1, 0),
             "SS_AMB vs Sed_AMB" = c(0, 0, 1, 0, -1, 0),
             "D1_AMB vs D1_HOT" = c(1, -1, 0, 0, 0, 0),
             "SS_AMB vs SS_HOT" = c(0, 0, 0, 0, 1, -1),
             "SED_AMB vs SED_HOT" = c(0, 0, 1, -1, 0, 0)
             )
```

```{r}
colour.brm2 |>
  emmeans(~symxtemp) |>
  contrast(list(symxtemp=cmat)) |>
  gather_emmeans_draws() |>
  dplyr::summarise(median_hdci(.value),
            Pl = mean(.value < 0),
            Pg = mean(.value > 0)
            )
```

```{r}
colour.dif.plot<- colour.brm2 %>%
  emmeans(~symxtemp) %>%
  gather_emmeans_draws() %>%
    ggplot() +
  #geom_vline(xintercept = 0, linetype = "dashed") +
  stat_slab(aes(
    x = .value, y = symxtemp,
    fill = stat(ggdist::cut_cdf_qi(cdf,
      .width = c(0.5, 0.8, 0.95),
      labels = scales::percent_format()
    ))
  ), color = "black") +
  scale_fill_brewer("Interval", direction = -1, na.translate = FALSE) +
  theme_classic() +
  labs(x = "Change in colour", y = "Treatment")
colour.dif.plot
```

Final Colour
form
```{r}
colour.form <- bf(T6_2 ~ symxtemp + (1|Plate.Slide),
                     family = gaussian()
                   )
```

priors
```{r fitModel2h1, results='markdown', eval=TRUE, hidden=TRUE, cache=TRUE}
priors<- prior("normal(20, 20)", class = "Intercept") +
    prior("normal(-10, 10)", class = "b") +
    #prior("normal(0, 0.5)", class = "sd", group = "Tank") +
    prior("normal(5, 10)", class = "sd", group = "Plate.Slide") +
    prior(student_t(3, 0, 2), class = 'sigma') +
    prior(cauchy(0,2.5), class = 'sd')
```

run model
```{r}
colour.brm3 <- brm(colour.form, 
                  data = colour,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 10000,
                  warmup = 5000,
                  chains = 4,
                  thin = 20,
                  control = list(adapt_delta = 0.99, max_treedepth = 20),
                  refresh = 0
                  )
```

View prior and posterior
```{r}
colour.brm3 %>% SUYR_prior_and_posterior()
```

trace plots
```{r}
colour.brm3$fit |> stan_trace()
```

Autocorrelation
```{r}
colour.brm3$fit |> stan_ac()
```

Rhat
```{r}
colour.brm3$fit |> stan_rhat()
```

Effective sample size
```{r}
colour.brm3$fit |> stan_ess()
```

Density overlay
```{r modelValidation5b, results='markdown', eval=TRUE, hidden=TRUE, fig.width=6, fig.height=4}
colour.brm3 %>% pp_check(type = 'dens_overlay', nsamples = 250)
```

DHARMa residuals
```{r}
colour.resids<- make_brms_dharma_res(colour.brm3, integerResponse = FALSE)
wrap_elements(~testUniformity(colour.resids)) +
wrap_elements(~plotResiduals(colour.resids, form = factor(rep(1,nrow(colour))))) +
wrap_elements(~plotResiduals(colour.resids, quantreg = FALSE))+
wrap_elements(~testDispersion(colour.resids))
```

Summarise model
conditional effects
```{r partialPlot2d, results='markdown', eval=TRUE, hidden=TRUE, fig.width=8, fig.height=5}
colour.brm3 %>%
    conditional_effects("symxtemp") %>%
    plot(points = TRUE)
```

```{r}
colour.brm3 |>
  emmeans(~ symxtemp) |>
  gather_emmeans_draws() |>
  dplyr::group_by(symxtemp) |>
  dplyr::summarise(
    median = median(.value),
    lower = quantile(.value, 0.025),
    upper = quantile(.value, 0.975)
  )
```

planned contrasts- compare situations
                 D1 vs Sed       D1 vs SS        SS vs Sed       Central vs Hybrid       
D1_AMB     |      0                -1/2                0                   -1/3
D1_HOT     |     -1                -1/2                0                    1/4
Jewell_AMB |      0              0                  1/2                 -1/3
Jewell_HOT |      0              0                  1/2                  1/4
SS_AMB     |      0                1/2               -1/2                 -1/3
SS_HOT     |      1                1/2               -1/2                  1/4

```{r}
cmat<- cbind(
             "D1_HOT vs Sed_HOT"=c(0, -1, 0, 1, 0, 0),
             "D1_HOT vs SS_HOT" = c(0, -1, 0, 0, 0, 1),
             "SS_HOT vs Sed_HOT" = c(0, 0, 0, 1, 0, -1),
             "D1_AMB vs Sed_AMB"=c(-1, 0, 1, 0, 0, 0),
             "D1_AMB vs SS_AMB" = c(-1, 0, 0, 0, 1, 0),
             "SS_AMB vs Sed_AMB" = c(0, 0, 1, 0, -1, 0),
             "D1_AMB vs D1_HOT" = c(1, -1, 0, 0, 0, 0),
             "SS_AMB vs SS_HOT" = c(0, 0, 0, 0, 1, -1),
             "SED_AMB vs SED_HOT" = c(0, 0, 1, -1, 0, 0)
             )
```
 
```{r}
colour.brm3 |>
  emmeans(~symxtemp) |>
  contrast(list(symxtemp=cmat)) |>
  gather_emmeans_draws() |>
  dplyr::summarise(median_hdci(.value),
            Pl = mean(.value < 0),
            Pg = mean(.value > 0)
            )
```

plot
```{r}
colour.plot<- colour.brm3 %>%
  emmeans(~symxtemp) %>%
  gather_emmeans_draws() %>%
    ggplot() +
  #geom_vline(xintercept = 0, linetype = "dashed") +
  stat_slab(aes(
    x = .value, y = symxtemp,
    fill = stat(ggdist::cut_cdf_qi(cdf,
      .width = c(0.5, 0.8, 0.95),
      labels = scales::percent_format()
    ))
  ), color = "black") +
  scale_fill_brewer("Interval", direction = -1, na.translate = FALSE) +
  theme_classic() +
  labs(x = "Final colour", y = "Treatment")
colour.plot
#saveRDS(colour.plot, file = "colour.plot_A.tenuis_juv.rds")
```

########
Read in the data
```{r readData, results='markdown', eval=TRUE}
yii <- read.csv("/Users/alexmacadam/Documents/GitHub/AGF-Physiology/data/Juvenile-PAM-2020.csv", header=T, fileEncoding="UTF-8-BOM") |>
  mutate(Family = factor(Family),
         Species = factor(Species),
         Temp = factor(Temp),
         Tank = factor(Tank),
         Timepoint = factor(Timepoint),
         Cassette = factor(Cassette),
         Plate.Slide = factor(Plate.Slide),
         Treatment = factor(Treatment),
         Cone = factor(Cone),
         Crosstype = factor(Crosstype),
         popcross = factor(popcross),
         SINGLE.CLUMP = factor(SINGLE.CLUMP),
         Mom = factor(Mom),
         Dad = factor(Dad)
         ) |>
  filter(Crosstype == "IndivCross", SINGLE.CLUMP == "S", Timepoint %in% c("T-2","T6"))

#Add Temp x Treatment column
yii$symxtemp <- paste(yii$Treatment, yii$Temp, sep='_')
yii$symxtemp <- as.factor(yii$symxtemp)
```

```{r}
#reshape data
yii2<- yii |>
  dplyr::select(Treatment, Timepoint, Temp,  YII_micro, IndID, popcross, Tank, Cassette, Plate.Slide, symxtemp) |>
  mutate(popcross = factor(popcross)) |>
  pivot_wider(names_from = Timepoint, values_from = YII_micro) |>
  tidyr::unnest()

yii2<- yii2 |>
  mutate(YII_change = (`T6` - `T-2`)*1000) |> #times 1000 to make bigger numbers
  drop_na(YII_change)

yii2$`T-2`[yii2$`T-2`<0.001] <- 0.001
yii2$T6[yii2$T6<0.001] <- 0.001

yii2<- yii2 |>
  mutate(YII_p_change = ((`T6` - `T-2`)/`T-2`)*100) |>
  drop_na(YII_p_change)
```

```{r}
#create final YII
final_yii<- yii |>
  filter(Timepoint == "T6") |>
  mutate(YII_micro = YII_micro * 100)
```

form
```{r}
yii.form <- bf(YII_change ~ symxtemp + (1|Plate.Slide),
                     family = gaussian()
                   )
```


```{r fitModel2h1, results='markdown', eval=TRUE, hidden=TRUE, cache=TRUE}
priors<- prior("normal(0, 200)", class = "Intercept") +
    prior("normal(0, 200)", class = "b") +
    #prior("normal(0, 0.5)", class = "sd", group = "Tank") +
    prior("normal(0, 300)", class = "sd", group = "Plate.Slide") +
    prior(student_t(3, 0, 100), class = 'sigma') +
    prior(cauchy(0,2.5), class = 'sd')
```

```{r}
YII.brm3 <- brm(yii.form, 
                  data = yii2,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 10000,
                  warmup = 5000,
                  chains = 4,
                  thin = 20,
                  control = list(adapt_delta = 0.99, max_treedepth = 20),
                  refresh = 0
                  )
```

View prior and posterior
```{r}
YII.brm3 %>% SUYR_prior_and_posterior()
```

Trace plots
```{r}
YII.brm3$fit |> stan_trace()
```

Autocorrelation
```{r}
YII.brm3$fit |> stan_ac()
```

Rhat
```{r}
YII.brm3$fit |> stan_rhat()
```

Effective sample size
```{r}
YII.brm3$fit |> stan_ess()
```

Density overlay
```{r modelValidation5b, results='markdown', eval=TRUE, hidden=TRUE, fig.width=6, fig.height=4}
YII.brm3 %>% pp_check(type = 'dens_overlay', nsamples = 250)
```

DHARMa residuals
```{r}
YII.resids<- make_brms_dharma_res(YII.brm3, integerResponse = FALSE)
wrap_elements(~testUniformity(YII.resids)) +
wrap_elements(~plotResiduals(YII.resids, form = factor(rep(1,nrow(YII))))) +
wrap_elements(~plotResiduals(YII.resids, quantreg = FALSE))+
wrap_elements(~testDispersion(YII.resids))
```

conditional effects
```{r partialPlot2d, results='markdown', eval=TRUE, hidden=TRUE, fig.width=8, fig.height=5}
YII.brm3 %>%
    conditional_effects("symxtemp") %>%
    plot(points = TRUE)
```

planned contrasts- compare situations
                 D1 vs Sed       D1 vs SS        SS vs Sed       Central vs Hybrid       
D1_AMB     |      0                -1/2                0                   -1/3
D1_HOT     |     -1                -1/2                0                    1/4
Jewell_AMB |      0              0                  1/2                 -1/3
Jewell_HOT |      0              0                  1/2                  1/4
SS_AMB     |      0                1/2               -1/2                 -1/3
SS_HOT     |      1                1/2               -1/2                  1/4
```{r}
cmat<- cbind(
             "D1_HOT vs Sed_HOT"=c(0, -1, 0, 1, 0, 0),
             "D1_HOT vs SS_HOT" = c(0, -1, 0, 0, 0, 1),
             "SS_HOT vs Sed_HOT" = c(0, 0, 0, 1, 0, -1),
             "D1_AMB vs Sed_AMB"=c(-1, 0, 1, 0, 0, 0),
             "D1_AMB vs SS_AMB" = c(-1, 0, 0, 0, 1, 0),
             "SS_AMB vs Sed_AMB" = c(0, 0, 1, 0, -1, 0),
             "D1_AMB vs D1_HOT" = c(1, -1, 0, 0, 0, 0),
             "SS_AMB vs SS_HOT" = c(0, 0, 0, 0, 1, -1),
             "SED_AMB vs SED_HOT" = c(0, 0, 1, -1, 0, 0)
             )
```

```{r}
YII.brm3 |>
  emmeans(~symxtemp) |>
  contrast(list(symxtemp=cmat)) |>
  gather_emmeans_draws() |>
  dplyr::summarise(median_hdci(.value),
            Pl = mean(.value < 0),
            Pg = mean(.value > 0)
            )
```

```{r}
YII.change.plot<- YII.brm3 %>%
  emmeans(~symxtemp) %>%
  gather_emmeans_draws() %>%
    ggplot() +
  #geom_vline(xintercept = 0, linetype = "dashed") +
  stat_slab(aes(
    x = .value, y = symxtemp,
    fill = stat(ggdist::cut_cdf_qi(cdf,
      .width = c(0.5, 0.8, 0.95),
      labels = scales::percent_format()
    ))
  ), color = "black") +
  scale_fill_brewer("Interval", direction = -1, na.translate = FALSE) +
  theme_classic() +
  labs(x = "Change in YII", y = "Treatment")
YII.change.plot
```

#####################################################################
Final YII

form
```{r}
yii.form <- bf(YII_micro ~ symxtemp + (1|Plate.Slide),
                     family = gaussian()
                   )
```

priors
```{r fitModel2h1, results='markdown', eval=TRUE, hidden=TRUE, cache=TRUE}
priors<- prior("normal(10, 20)", class = "Intercept") +
    prior("normal(0, 20)", class = "b") +
    #prior("normal(0, 0.5)", class = "sd", group = "Tank") +
    prior("normal(0, 20)", class = "sd", group = "Plate.Slide") +
    prior(student_t(3, 0, 20), class = 'sigma') +
    prior(cauchy(0,2.5), class = 'sd')
```

run model
```{r}
YII.brm2 <- brm(yii.form, 
                  data = yii,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 10000,
                  warmup = 5000,
                  chains = 4,
                  thin = 20,
                  control = list(adapt_delta = 0.99, max_treedepth = 20),
                  refresh = 0
                  )
```

View prior and posterior
```{r}
YII.brm2 %>% SUYR_prior_and_posterior()
```

Trace plots
```{r}
YII.brm2$fit |> stan_trace()
```

Autocorrelation
```{r}
YII.brm2$fit |> stan_ac()
```

Rhat
```{r}
YII.brm2$fit |> stan_rhat()
```

Effective sample size
```{r}
YII.brm2$fit |> stan_ess()
```

density overlay
```{r modelValidation5b, results='markdown', eval=TRUE, hidden=TRUE, fig.width=6, fig.height=4}
YII.brm2 %>% pp_check(type = 'dens_overlay', nsamples = 250)
```

DHARMa residuals
```{r}
YII.resids<- make_brms_dharma_res(YII.brm2, integerResponse = FALSE)
wrap_elements(~testUniformity(YII.resids)) +
wrap_elements(~plotResiduals(YII.resids, form = factor(rep(1,nrow(YII))))) +
wrap_elements(~plotResiduals(YII.resids, quantreg = FALSE))+
wrap_elements(~testDispersion(YII.resids))
```

Summarise model
conditional effects
```{r partialPlot2d, results='markdown', eval=TRUE, hidden=TRUE, fig.width=8, fig.height=5}
YII.brm2 %>%
    conditional_effects("symxtemp") %>%
    plot(points = TRUE)
```

```{r}
YII.brm2 |>
  emmeans(~ symxtemp) |>
  gather_emmeans_draws() |>
  dplyr::group_by(symxtemp) |>
  dplyr::summarise(
    median = median(.value),
    lower = quantile(.value, 0.025),
    upper = quantile(.value, 0.975)
  )
```

planned contrasts- compare situations
                 D1 vs Sed       D1 vs SS        SS vs Sed       Central vs Hybrid       
D1_AMB     |      0                -1/2                0                   -1/3
D1_HOT     |     -1                -1/2                0                    1/4
Jewell_AMB |      0              0                  1/2                 -1/3
Jewell_HOT |      0              0                  1/2                  1/4
SS_AMB     |      0                1/2               -1/2                 -1/3
SS_HOT     |      1                1/2               -1/2                  1/4
```{r}
cmat<- cbind(
             "D1_HOT vs Sed_HOT"=c(0, -1, 0, 1, 0, 0),
             "D1_HOT vs SS_HOT" = c(0, -1, 0, 0, 0, 1),
             "SS_HOT vs Sed_HOT" = c(0, 0, 0, 1, 0, -1),
             "D1_AMB vs Sed_AMB"=c(-1, 0, 1, 0, 0, 0),
             "D1_AMB vs SS_AMB" = c(-1, 0, 0, 0, 1, 0),
             "SS_AMB vs Sed_AMB" = c(0, 0, 1, 0, -1, 0),
             "D1_AMB vs D1_HOT" = c(1, -1, 0, 0, 0, 0),
             "SS_AMB vs SS_HOT" = c(0, 0, 0, 0, 1, -1),
             "SED_AMB vs SED_HOT" = c(0, 0, 1, -1, 0, 0)
             )
```
 
```{r}
YII.brm2 |>
  emmeans(~symxtemp) |>
  contrast(list(symxtemp=cmat)) |>
  gather_emmeans_draws() |>
  dplyr::summarise(median_hdci(.value),
            Pl = mean(.value < 0),
            Pg = mean(.value > 0)
            )
```

```{r}
YII.plot<- YII.brm2 %>%
  emmeans(~symxtemp) %>%
  gather_emmeans_draws() %>%
    ggplot() +
  #geom_vline(xintercept = 0, linetype = "dashed") +
  stat_slab(aes(
    x = .value, y = symxtemp,
    fill = stat(ggdist::cut_cdf_qi(cdf,
      .width = c(0.5, 0.8, 0.95),
      labels = scales::percent_format()
    ))
  ), color = "black") +
  scale_fill_brewer("Interval", direction = -1, na.translate = FALSE) +
  theme_classic() +
  labs(x = "Final YII", y = "Treatment")
YII.plot
```

###
```{r}
surv.plot
a<- growth.plot + size.plot + colour.dif.plot + colour.plot + YII.change.plot + YII.plot + plot_layout(ncol = 2, guides = 'collect') + plot_annotation(tag_levels = 'A')
```

