---
title: "RNAeq quality control"
output: html_document
date: "2023-06-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Set working directory
```{r}
setwd("/Users/jc828813/Dropbox/PhD/Chapter 4- RNA/Chapter 4- RNA- scripts/Chapter-4--Gene-expression")
```

Load packages
```{r}
library(tidyverse)
```

Make a folder for the 
```{r}
system("mkdir -p raw_data/aten_larvae_reads")
```

Load file with sample names
```{r}
raw_files <- read_tsv("raw_data/aten_larvae_raw_files.tsv", col_names = c("folder","file"))
```

This creates a vector of file paths which we use later to construct URLs for downloading and paths for saving
```{r}
paths <- raw_files %>% 
  filter(folder=="aten_larvae_reads") %>% 
  unite("path",folder,file,sep="/") %>% 
  pull(path)


for (path in paths){
  
  # Construct a URL for downloading
  url <- paste("https://s3-ap-southeast-2.amazonaws.com/bc3203/",path,sep = "")

  # Construct a relative path for saving the file
  dest <- paste("raw_data/",path,sep="")

  if ( !file.exists(dest)){
    download.file(url, destfile = dest)
  }
}
```

As a rough check to ensure that the files downloaded properly we could simply list the number of reads in each file.
Here is some bash code to check an individual file:
```{r}
system("gunzip -c raw_data/aten_larvae_reads/Plate3_G1_S189_R2_001.fastq.gz | grep '^+$' | wc -l")
```

This bash code checks all the files. It should print the name of each file followed by a number indicating the number of reads the file contains
```{r}
system("
for f in raw_data/aten_larvae_reads/*.gz;do 
  echo $(basename $f) $(gunzip -c $f | grep '^+$' | wc -l)
done
")
```

Unzip the genome
```{r}
system("gunzip aten_final_0.11.fasta.gz")
```

save the gene info in a table
```{r}
gene_info <- "raw_data/aten_0.11.maker_post_001.genes.gff.gz"
gene_info2 <- read.gff(gene_info)

#how many genes are described?
gene_info2 %>% filter(type == "gene") %>% count()

#plot showing the number of genes in each chromosome
gene_info2 %>% filter(type == "gene") %>%
  group_by(seqid) %>%
  summarise(num_genes = n()) %>%
  ggplot(aes(x=seqid, y=num_genes)) +
  geom_col() +
  labs(x = "Chromosome", y = "Number of genes")

#Make a plot showing the density of genes on chromosomes 1, 2 and 3 as a function of genomic coordinate
gene_info2 %>% 
  filter(type == "gene") %>%
  filter(seqid == c("Sc0000000", "Sc0000003", "Sc0000003")) %>%
  ggplot(aes(x=start)) +
  geom_histogram() +
  facet_wrap(~seqid)

#Plot the distribution (frequency histogram or density plot) of gene lengths. Use a log scale for the x axis.
gene_info2 %>%
  filter(type == "gene") %>%
  mutate(length = end - start) %>%
  ggplot(aes(x=length)) +
  geom_histogram() +
  scale_x_log10()
```

```{r}
system("mkdir -p rsem")
```

```{r}
system("rsem-prepare-reference --gtf raw_data/aten_0.11.maker_post_001.genes.gff.gz --bowtie2 raw_data/aten_final_0.11.fasta rsem/aten_ref")
```


