---
title: "10.2_aten_juv_sPLS_combined"
output: html_document
date: "2024-08-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(limma)
library(Glimma)
library(edgeR)
library(tidyverse)
library(mixOmics) # import the mixOmics library
set.seed(5249) # for reproducibility, remove for normal use
```

#############
Host
Sample info
```{r}
coldata<- read.csv("/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Data/Gene_counts/read_info/aten_juvenile_read_info2.csv", row.names="ID") |> filter(Symbiont == "D1")
coldata <- coldata[,c("Cross","Temp","Symbiont","group")]
#convert columns to factors
coldata$Cross <- factor(coldata$Cross)
coldata$group <- factor(coldata$group)
coldata$Temp <- factor(coldata$Temp)
coldata$Symbiont <- factor(coldata$Symbiont)

#make sample lists for each symbiont treatment
coldata2<- coldata |> rownames_to_column(var = "Samples")
D1samples<- coldata2 |> dplyr::filter(Symbiont == "D1")
D1samples<- D1samples$Samples
```

Get Count Data
```{r}
x<- as.matrix(read.csv("/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Data/Gene_counts/count_tables/Juvenile_host/all/aten_juveniles_host_all_counts2.csv", row.names = 1))
#view data
#head(x,2)
rownames
rownames(coldata) <- sub("fb","", rownames(coldata))
#Do row names in coldata data match to column data in x
all(rownames(coldata) %in% colnames(x))
#are they in the same order?
all(rownames(coldata) == colnames(x))
#change order
x <- x[, rownames(coldata)]
#double check order
all(rownames(coldata) == colnames(x))
```

```{r}
#Create the matrix
x <- DGEList(counts = x)

# Assuming `sample_info` has a column named `group` for conditions
x$samples$group <- coldata$group

class(x)

dim(x)
```

```{r}
samplenames <- substring(colnames(x), 1, nchar(colnames(x)))
samplenames

colnames(x) <- samplenames
group <- as.factor(c("AmbientD1", "AmbientD1", "AmbientD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "HotD1", "AmbientD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "HotD1", "HotD1"))
x$samples$group <- group
```

```{r}
geneid <- rownames(x)
```

####
Data pre-processing

###
Transformations from the raw-scale
```{r}
cpm <- cpm(x)
lcpm <- cpm(x, log=TRUE)
```

```{r}
L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6
c(L, M)
summary(lcpm)
```

###
Removing genes that are lowly expressed
```{r}
table(rowSums(x$counts==0)==9)

keep.exprs <- filterByExpr(x, group=group)
x <- x[keep.exprs,, keep.lib.sizes=FALSE]
dim(x)
```

```{r}
lcpm.cutoff <- log2(10/M + 2/L)
library(RColorBrewer)
nsamples <- ncol(x)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
lcpm <- cpm(x, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
```

###
Normalising gene expression distributions
```{r}
x <- calcNormFactors(x, method = "TMM")
x$samples$norm.factors
x<- data.matrix(x)
x<- DESeq2::varianceStabilizingTransformation(x)
x<- t(x)
```

```{r}
y<-group
summary(y) # check the distribution of class labels
```

```{r}
x_D1<- x
y_d1<- y
```

##################
Symbionts
Get Count Data
```{r}
x<- as.matrix(read.csv("/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Data/Gene_counts/count_tables/Juvenile_symbiont/aten_juveniles_symb_Dtre_counts.csv", row.names = 1))
#x<- as.matrix(read.csv("/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Data/Gene_counts/count_tables/Juvenile_symbiont/aten_#juveniles_symb_Dtre_counts(outlier_filtered).csv", row.names = 1))
#view data
head(x,2)
rownames
rownames(coldata) <- sub("fb","", rownames(coldata))
#Do row names in coldata data match to column data in x
all(rownames(coldata) %in% colnames(x))
#are they in the same order?
all(rownames(coldata) == colnames(x))
#change order
x <- x[, rownames(coldata)]
#double check order
all(rownames(coldata) == colnames(x))
```

```{r}
#Create the matrix
x <- DGEList(counts = x)

# Assuming `sample_info` has a column named `group` for conditions
x$samples$group <- coldata$group
x$samples$Cross <- coldata$Cross

class(x)

dim(x)
```

```{r}
samplenames <- substring(colnames(x), 1, nchar(colnames(x)))
samplenames

colnames(x) <- samplenames
group <- as.factor(c("AmbientD1", "AmbientD1", "AmbientD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "HotD1", "AmbientD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "HotD1", "HotD1"))
#group <- as.factor(c("AmbientD1", "AmbientD1", "AmbientD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "AmbientD1", #"HotD1", "AmbientD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "AmbientD1", #"HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "HotD1", "AmbientD1", #"AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "HotD1"))
x$samples$group <- group

Cross <- as.factor(c("J8xJ2","J8xJ2","J8xD6","J8xJ3","J8xD13","J8xD13","J8xD13","J8xJ3","J8xJ3","J8xD13","J8xD6","J8xJ2","J8xD1","J8xD6","J8xJ2","J8xD1","J8xJ3","J8xD13","J8xD1","J8xD6","J8xJ2","J8xD1","J8xD6","J8xJ3","J8xD1","J8xJ2","J8xD6","J8xJ3","J8xD13","J8xD1","J8xP5","J8xP5","J8xP5","J8xP5","J8xP5","J8xP5"))
x$samples$Cross <- Cross
```

```{r}
geneid <- rownames(x)
```

####
Data pre-processing

###
Transformations from the raw-scale
```{r}
cpm <- cpm(x)
lcpm <- cpm(x, log=TRUE)
```

```{r}
L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6
c(L, M)
summary(lcpm)
```

###
Removing genes that are lowly expressed
```{r}
table(rowSums(x$counts==0)==9)

keep.exprs <- filterByExpr(x, group=group)
x <- x[keep.exprs,, keep.lib.sizes=FALSE]
dim(x)
```

```{r}
lcpm.cutoff <- log2(10/M + 2/L)
library(RColorBrewer)
nsamples <- ncol(x)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
lcpm <- cpm(x, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
```

###
Normalising gene expression distributions
```{r}
x <- calcNormFactors(x, method = "TMM")
x$samples$norm.factors
x<- data.matrix(x)
x<- DESeq2::varianceStabilizingTransformation(x)
x<- t(x)
```

```{r}
y<-group
summary(y) # check the distribution of class labels
```

```{r}
x_Dtre<- x
y_Dtre<- y
```

###############
Preliminary Analysis with PCA
```{r}
pca.gene <- pca(x_D1, ncomp = 10, center = TRUE, scale = TRUE)
pca.clinic <- pca(x_Dtre, ncomp = 10, center = TRUE, scale = TRUE)

plot(pca.gene)
plot(pca.clinic)
```

```{r}
plotIndiv(pca.gene, comp = c(1, 2), 
          group = y, 
          ind.names = y, 
          legend = TRUE, 
          title = 'Liver gene, PCA comp 1 - 2')

plotIndiv(pca.clinic, comp = c(1, 2), 
          group = y, 
          ind.names = y, 
          legend = TRUE, title = 'Liver clinic, PCA comp 1 - 2')
```

Initial sPLS model
```{r}
spls.liver <- spls(X = x_D1, Y = x_Dtre, ncomp = 6, mode = 'regression')

#saveRDS(spls.liver, file = "/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Scripts/Codes for RNAseq analysis/10.2_aten_host_mixomics1.rds")
spls.liver<- readRDS("/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Scripts/Codes for RNAseq analysis/10.2_aten_host_mixomics1.rds")
```

Tuning sPLS
```{r}
# repeated CV tuning of component count
perf.spls.liver <- perf(spls.liver, validation = 'Mfold',
                        folds = 3, nrepeat = 50,
                        progressBar = TRUE)

#saveRDS(perf.spls.liver, file = "/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Scripts/Codes for RNAseq analysis/10.2_aten_host_mixomics2.rds")
perf.spls.liver<- readRDS("/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Scripts/Codes for RNAseq analysis/10.2_aten_host_mixomics2.rds")

plot(perf.spls.liver, criterion = 'Q2.total')
```


Selecting the number of variables
```{r}
# set range of test values for number of variables to use from X dataframe
list.keepX <- c(seq(20, 50, 5))
# set range of test values for number of variables to use from Y dataframe
list.keepY <- c(3:10) 


tune.spls.liver <- tune.spls(x_D1, x_Dtre, ncomp = 2,
                              test.keepX = list.keepX,
                              test.keepY = list.keepY,
                              nrepeat = 1, folds = 10, # use 10 folds
                              mode = 'regression', measure = 'cor') 

#saveRDS(tune.spls.liver, file = "/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Scripts/Codes for RNAseq analysis/10.2_aten_host_mixomics3.rds")
tune.spls.liver<- readRDS("/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Scripts/Codes for RNAseq analysis/10.2_aten_host_mixomics3.rds")

plot(tune.spls.liver)         # use the correlation measure for tuning
```

```{r}
tune.spls.liver$choice.keepX
```

```{r}
tune.spls.liver$choice.keepY
```

```{r}
# extract optimal number of variables for X dataframe
optimal.keepX <- tune.spls.liver$choice.keepX 

# extract optimal number of variables for Y datafram
optimal.keepY <- tune.spls.liver$choice.keepY

optimal.ncomp <-  length(optimal.keepX) # extract optimal number of components
```

Final Model
```{r}
# use all tuned values from above
final.spls.liver <- spls(X, Y, ncomp = optimal.ncomp, 
                    keepX = optimal.keepX,
                    keepY = optimal.keepY,
                    mode = "regression") # explanitory approach being used, 
                                         # hence use regression mode
```

Plots
```{r}
plotIndiv(final.spls.liver, ind.names = FALSE, 
         rep.space = "X-variate", # plot in X-variate subspace
         group = liver.toxicity$treatment$Time.Group, # colour by time group
         pch = as.factor(liver.toxicity$treatment$Dose.Group), 
         col.per.group = color.mixo(1:4), 
         legend = TRUE, legend.title = 'Time', legend.title.pch = 'Dose')

plotIndiv(final.spls.liver, ind.names = FALSE,
         rep.space = "Y-variate", # plot in Y-variate subspace
         group = liver.toxicity$treatment$Time.Group, # colour by time group
         pch = as.factor(liver.toxicity$treatment$Dose.Group), 
         col.per.group = color.mixo(1:4), 
         legend = TRUE, legend.title = 'Time', legend.title.pch = 'Dose')
```

```{r}
plotIndiv(final.spls.liver, ind.names = FALSE, 
         rep.space = "XY-variate", # plot in averaged subspace
         group = liver.toxicity$treatment$Time.Group, # colour by time group
         pch = as.factor(liver.toxicity$treatment$Dose.Group), # select symbol
         col.per.group = color.mixo(1:4),                      # by dose group
         legend = TRUE, legend.title = 'Time', legend.title.pch = 'Dose')
```

```{r}
col.tox <- color.mixo(as.numeric(as.factor(liver.toxicity$treatment[, 4]))) # create set of colours
plotIndiv(final.spls.liver, ind.names = FALSE, 
          rep.space = "XY-variate", # plot in averaged subspace
          axes.box = "both", col = col.tox, style = '3d')
```

```{r}
plotArrow(final.spls.liver, ind.names = FALSE,
          group = liver.toxicity$treatment$Time.Group, # colour by time group
          col.per.group = color.mixo(1:4),
          legend.title = 'Time.Group')
```

```{r}
# form new perf() object which utilises the final model
perf.spls.liver <- perf(final.spls.liver, 
                          folds = 5, nrepeat = 10, # use repeated cross-validation
                          validation = "Mfold", 
                          dist = "max.dist",  # use max.dist measure
                          progressBar = FALSE)

# plot the stability of each feature for the first two components, 
# 'h' type refers to histogram
par(mfrow=c(1,2)) 
plot(perf.spls.liver$features$stability.X[[1]], type = 'h',
     ylab = 'Stability',
     xlab = 'Features',
     main = '(a) Comp 1', las =2,
     xlim = c(0, 150))
plot(perf.spls.liver$features$stability.X$comp2, type = 'h',
     ylab = 'Stability',
     xlab = 'Features',
     main = '(b) Comp 2', las =2,
     xlim = c(0, 300))
```

```{r}
plotVar(final.spls.liver, cex = c(3,4), var.names = c(FALSE, TRUE))
```

```{r}
color.edge <- color.GreenRed(50)  # set the colours of the connecting lines

# X11() # To open a new window for Rstudio
network(final.spls.liver, comp = 1:2,
        cutoff = 0.7, # only show connections with a correlation above 0.7
        shape.node = c("rectangle", "circle"),
        color.node = c("cyan", "pink"),
        color.edge = color.edge,
        save = 'png', # save as a png to the current working directory
        name.save = 'sPLS Liver Toxicity Case Study Network Plot')
```
