---
title: "7.2.2_aten_juv_limma_amb"
output: html_document
date: "2024-08-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(limma)
library(Glimma)
library(edgeR)
library(tidyverse)
set.seed(123)
```

Sample info
```{r}
coldata<- read.csv("/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Data/Gene_counts/read_info/aten_juvenile_read_info2.csv", row.names="ID") |>
  filter(Temp == "Ambient")
coldata <- coldata[,c("Cross","Temp","Symbiont","group")]
#convert columns to factors
coldata$Cross <- factor(coldata$Cross)
coldata$group <- factor(coldata$group)
coldata$Temp <- factor(coldata$Temp)
coldata$Symbiont <- factor(coldata$Symbiont)

#make sample lists for each symbiont treatment
coldata2<- coldata |> rownames_to_column(var = "Samples")
D1samples<- coldata2 |> dplyr::filter(Symbiont == "D1")
SSsamples<- coldata2 |> dplyr::filter(Symbiont == "SS")
Wildsamples<-  coldata2 |> dplyr::filter(Symbiont == "Wild")
D1samples<- D1samples$Samples
SSsamples<- SSsamples$Samples
Wildsamples<- Wildsamples$Samples
samples<- c(D1samples, SSsamples, Wildsamples)
```

Get Count Data
```{r}
x<- as.matrix(read.csv("/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Data/Gene_counts/count_tables/Juvenile_host/all/aten_juveniles_host_all_counts2.csv", row.names = 1))
# Keep only Ambient samples
x <- x[, colnames(x) %in% samples]

#view data
#head(x,2)
rownames
rownames(coldata) <- sub("fb","", rownames(coldata))
#Do row names in coldata data match to column data in x
all(rownames(coldata) %in% colnames(x))
#are they in the same order?
all(rownames(coldata) == colnames(x))
#change order
x <- x[, rownames(coldata)]
#double check order
all(rownames(coldata) == colnames(x))
```

```{r}
#Create the matrix
x <- DGEList(counts = x)

# Assuming `sample_info` has a column named `group` for conditions
x$samples$group <- coldata$group

class(x)

dim(x)
```

```{r}
samplenames <- substring(colnames(x), 1, nchar(colnames(x)))
samplenames

colnames(x) <- samplenames

group <- as.factor(c("AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientWild","AmbientWild","AmbientWild","AmbientWild","AmbientWild","AmbientWild","AmbientWild","AmbientWild","AmbientWild","AmbientWild","AmbientWild","AmbientWild","AmbientWild","AmbientWild"))
x$samples$group <- group

Cross <- as.factor(c("J8xJ2","J8xJ2","J8xD6","J8xJ3","J8xD13","J8xD13","J8xJ3","J8xD13","J8xD6","J8xD6","J8xJ3","J8xD1","J8xD1","J8xJ2","J8xD1","J8xP5","J8xP5","J8xP5","J8xJ3","J8xJ2","J8xD6","J8xJ3","J8xD1","J8xJ2","J8xJ2","J8xD13","J8xJ3","J8xD1","J8xD6","J8xD6","J8xD13","J8xD6","J8xD1","J8xD13","J8xP5","J8xP5","J8xD1","J8xD13","J8xP5","J8xJ3","J8xJ2","J8xJ2","J8xP5","J8xJ2","J8xD6","J8xD6","J8xD6","J8xJ3","J8xP5","J8xJ3"))

x$samples$Cross <- Cross
```

```{r}
geneid <- rownames(x)
```

####
Data pre-processing

###
Transformations from the raw-scale
```{r}
cpm <- cpm(x)
lcpm <- cpm(x, log=TRUE)
```

```{r}
L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6
c(L, M)
summary(lcpm)
```

###
Removing genes that are lowly expressed
```{r}
table(rowSums(x$counts==0)==9)

keep.exprs <- filterByExpr(x, group=group)
x <- x[keep.exprs,, keep.lib.sizes=FALSE]
dim(x)
```

```{r}
lcpm.cutoff <- log2(10/M + 2/L)
library(RColorBrewer)
nsamples <- ncol(x)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
lcpm <- cpm(x, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
```

###
Normalising gene expression distributions
```{r}
x <- calcNormFactors(x, method = "TMM")
x$samples$norm.factors
```

```{r}
x2 <- x
x2$samples$norm.factors <- 1
x2$counts[,1] <- ceiling(x2$counts[,1]*0.05)
x2$counts[,2] <- x2$counts[,2]*5
```

```{r}
par(mfrow=c(1,2))
lcpm <- cpm(x2, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="A. Example: Unnormalised data",ylab="Log-cpm")
x2 <- calcNormFactors(x2)  
x2$samples$norm.factors
```

```{r}
lcpm <- cpm(x2, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="B. Example: Normalised data",ylab="Log-cpm")
```

###
Unsupervised clustering of samples
```{r}
lcpm <- cpm(x, log=TRUE)
par(mfrow=c(1,2))
col.group <- group
levels(col.group) <-  brewer.pal(nlevels(col.group), "Set1")
col.group <- as.character(col.group)
plotMDS(lcpm, labels=group, col=col.group)
title(main="A. Sample groups")
```

####
Differential expression analysis

###
Creating a design matrix and contrasts
```{r}
design <- model.matrix(~0+group+Cross)
colnames(design) <- gsub("group", "", colnames(design))
design
```

```{r}
contr.matrix <- makeContrasts(
   AmbD1vsAmbSS = AmbientD1-AmbientSS, 
   AmbD1vsAmbWild = AmbientD1-AmbientWild, 
   AmbSSvsAmbWild = AmbientSS-AmbientWild,
   
   levels = colnames(design))
contr.matrix
```

Removing heteroscedascity from count data
```{r}
par(mfrow=c(1,2))
v <- voom(x, design, plot=TRUE)
v
```

```{r}
vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=contr.matrix)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")
```

```{r}
#summary(decideTests(efit))
et <- decideTests(efit)
summary(et)
```

```{r}
#most strict filtering: log-FC that is significantly greater than 1 (equivalent to a 2-fold difference)
#tfit <- treat(vfit, lfc=1)
#dt <- decideTests(tfit)
#summary(dt)

#de.common <- which(dt[,1]!=0 & dt[,2]!=0)
#length(de.common)

#head(tfit$genes$SYMBOL[de.common], n=20)
```

Venn diagram of DEGs
```{r}
vennDiagram(et[,1:3], circle.col=c("turquoise","darkgoldenrod1", "salmon"))
```

Heatmaps
make gene list
```{r}
lcpm2<- lcpm |> as.data.frame() |> rownames_to_column(var = "Genes")

#select only genes
genes<- lcpm2 |> dplyr::select(Genes)
```

###############
D1
Examining individual DE genes from top to bottom
```{r}
AmbD1.vs.AmbSS <- topTreat(efit, coef=1, n=Inf)
AmbD1.vs.AmbWild <- topTreat(efit, coef=2, n=Inf)
AmbSS.vs.AmbWild <- topTreat(efit, coef=3, n=Inf)
head(AmbD1.vs.AmbSS)
```

```{r}
AmbD1.vs.AmbSS<- AmbD1.vs.AmbSS |> rownames_to_column(var = "gene") |> dplyr::select(gene, logFC, adj.P.Val) |> dplyr::rename(lfc = logFC, p = adj.P.Val)
#write.csv(AmbD1.vs.AmbSS, "/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Data/KOG/AmbD1.vs.AmbSS.csv")

AmbD1.vs.AmbWild<- AmbD1.vs.AmbWild |> rownames_to_column(var = "gene") |> dplyr::select(gene, logFC, adj.P.Val) |> dplyr::rename(lfc = logFC, p = adj.P.Val)
#write.csv(AmbD1.vs.AmbWild, "/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Data/KOG/AmbD1.vs.AmbWild.csv")

AmbSS.vs.AmbWild<- AmbSS.vs.AmbWild |> rownames_to_column(var = "gene") |> dplyr::select(gene, logFC, adj.P.Val) |> dplyr::rename(lfc = logFC, p = adj.P.Val)
#write.csv(AmbSS.vs.AmbWild, "/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Data/KOG/AmbSS.vs.AmbWild.csv")
```

Useful graphical representations of differential expression results
```{r}
#D1 plot
plotMD(efit, column=1, status=et[,1], main=colnames(efit)[1], 
       xlim=c(-8,13))
```

Heatmaps
```{r}
#filter lcpm2 for only samples with each symbiont
#D1
D1_lcpm<- lcpm2 |> dplyr::select(matches(D1samples))
D1_lcpm<- cbind(genes, D1_lcpm)
D1_lcpm2<- column_to_rownames(D1_lcpm, var = "Genes")
D1_lcpm2<- data.matrix(D1_lcpm2)
```

```{r}
#D1 heatmap
library(gplots)
AmbD1.vs.AmbSS.2<-AmbD1.vs.AmbSS |> rownames_to_column(var = "Genes")
AmbD1.vs.AmbSS.topgenes <- AmbD1.vs.AmbSS.2$Genes[1:100]

#i <- which(v$genes$Genes %in% AmbD1.vs.HotD1.topgenes)
i_D1 <- which(D1_lcpm$Genes %in% AmbD1.vs.AmbSS.topgenes)
mycol <- colorpanel(1000,"blue","white","red")

#filter top 100 genes in matrix
D1_lcpm_100<- D1_lcpm |> dplyr::filter(Genes %in% AmbD1.vs.AmbSS.topgenes)
D1_lcpm_100<- column_to_rownames(D1_lcpm_100, var = "Genes")
D1_lcpm_100<- data.matrix(D1_lcpm_100)

groupD1 <- as.factor(c("AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientD1","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS","AmbientSS"))

heatmap.2(D1_lcpm_100,
   scale="row",
   #labRow=v$genes$SYMBOL[i], 
   labCol=groupD1,
   labRow="none",
   col=mycol, 
   trace="none", 
   density.info="none",
   #margin=c(8,6), 
   #lhei=c(2,10), 
   dendrogram="column"
   )
```

###############
SS
Examining individual DE genes from top to bottom
```{r}
AmbSS.vs.HotSS <- topTreat(efit, coef=2, n=Inf)
head(AmbSS.vs.HotSS)
```

Useful graphical representations of differential expression results
```{r}
#SS plot
plotMD(efit, column=2, status=et[,2], main=colnames(efit)[2], 
       xlim=c(-8,13))
```

Heatmaps
```{r}
#filter lcpm2 for only samples with each symbiont
#SS
SS_lcpm<- lcpm2 |> dplyr::select(matches(SSsamples))
SS_lcpm<- cbind(genes, SS_lcpm)
SS_lcpm2<- column_to_rownames(SS_lcpm, var = "Genes")
SS_lcpm2<- data.matrix(SS_lcpm2)
```

```{r}
#SS heatmap
library(gplots)
AmbSS.vs.HotSS.2<-AmbSS.vs.HotSS |> rownames_to_column(var = "Genes")
AmbSS.vs.HotSS.topgenes <- AmbSS.vs.HotSS.2$Genes[1:100]

#i <- which(v$genes$Genes %in% AmbSS.vs.HotSS.topgenes)
i_SS <- which(SS_lcpm$Genes %in% AmbSS.vs.HotSS.topgenes)
mycol <- colorpanel(1000,"blue","white","red")

#filter top 100 genes in matrix
SS_lcpm_100<- SS_lcpm |> dplyr::filter(Genes %in% AmbSS.vs.HotSS.topgenes)
SS_lcpm_100<- column_to_rownames(SS_lcpm_100, var = "Genes")
SS_lcpm_100<- data.matrix(SS_lcpm_100)

groupSS <- as.factor(c("AmbientSS", "HotSS", "HotSS", "AmbientSS", "AmbientSS", "HotSS", "AmbientSS", "AmbientSS", "AmbientSS", "AmbientSS", "HotSS", "AmbientSS", "HotSS", "AmbientSS", "AmbientSS", "HotSS", "HotSS", "HotSS", "AmbientSS", "AmbientSS", "HotSS", "AmbientSS", "HotSS", "AmbientSS", "HotSS", "HotSS", "AmbientSS", "AmbientSS", "HotSS", "HotSS", "AmbientSS", "AmbientSS"))

heatmap.2(SS_lcpm_100, 
   scale="row",
   #labRow=v$genes$SYMBOL[i], 
   labCol=groupSS,
   labRow="none",
   col=mycol, 
   trace="none", 
   density.info="none",
   #margin=c(8,6), 
   #lhei=c(2,10), 
   dendrogram="column"
   )
```

###########
Wild
Examining individual DE genes from top to bottom
```{r}
AmbWild.vs.HotWild <- topTreat(efit, coef=3, n=Inf)
head(AmbWild.vs.HotWild)
```

Useful graphical representations of differential expreWildion results
```{r}
#Wild plot
plotMD(efit, column=3, status=et[,3], main=colnames(efit)[3], 
       xlim=c(-8,13))
```

Heatmaps
```{r}
#filter lcpm2 for only samples with each symbiont
#Wild
Wild_lcpm<- lcpm2 |> dplyr::select(matches(Wildsamples))
Wild_lcpm<- cbind(genes, Wild_lcpm)
Wild_lcpm2<- column_to_rownames(Wild_lcpm, var = "Genes")
Wild_lcpm2<- data.matrix(Wild_lcpm2)
```

```{r}
#Wild heatmap
library(gplots)
AmbWild.vs.HotWild.2<-AmbWild.vs.HotWild |> rownames_to_column(var = "Genes")
AmbWild.vs.HotWild.topgenes <- AmbWild.vs.HotWild.2$Genes[1:100]

#i <- which(v$genes$Genes %in% AmbWild.vs.HotWild.topgenes)
i_Wild <- which(Wild_lcpm$Genes %in% AmbWild.vs.HotWild.topgenes)
mycol <- colorpanel(1000,"blue","white","red")

#filter top 100 genes in matrix
Wild_lcpm_100<- Wild_lcpm |> dplyr::filter(Genes %in% AmbWild.vs.HotWild.topgenes)
Wild_lcpm_100<- column_to_rownames(Wild_lcpm_100, var = "Genes")
Wild_lcpm_100<- data.matrix(Wild_lcpm_100)

groupWild <- as.factor(c("HotWild", "AmbientWild", "AmbientWild", "HotWild", "AmbientWild", "AmbientWild", "HotWild", "AmbientWild", "AmbientWild", "AmbientWild", "HotWild", "AmbientWild", "HotWild", "AmbientWild", "AmbientWild", "AmbientWild", "AmbientWild", "HotWild", "HotWild", "HotWild", "AmbientWild", "AmbientWild", "HotWild"))

heatmap.2(Wild_lcpm_100, 
   scale="row",
   #labRow=v$genes$SYMBOL[i], 
   labCol=groupWild,
   labRow="none",
   col=mycol, 
   trace="none", 
   density.info="none",
   #margin=c(8,6), 
   #lhei=c(2,10), 
   dendrogram="column"
   )
```
