---
title: "10.1_aten_juv_rCCA"
output: html_document
date: "2024-08-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(limma)
library(Glimma)
library(edgeR)
library(tidyverse)
library(mixOmics) # import the mixOmics library
set.seed(5249) # for reproducibility, remove for normal use
```

#############
Host
Sample info
```{r}
coldata<- read.csv("/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Data/Gene_counts/read_info/aten_juvenile_read_info2.csv", row.names="ID") |> filter(Symbiont == "D1")
coldata <- coldata[,c("Cross","Temp","Symbiont","group")]
#convert columns to factors
coldata$Cross <- factor(coldata$Cross)
coldata$group <- factor(coldata$group)
coldata$Temp <- factor(coldata$Temp)
coldata$Symbiont <- factor(coldata$Symbiont)

#make sample lists for each symbiont treatment
coldata2<- coldata |> rownames_to_column(var = "Samples")
D1samples<- coldata2 |> dplyr::filter(Symbiont == "D1")
D1samples<- D1samples$Samples
```

Get Count Data
```{r}
x<- as.matrix(read.csv("/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Data/Gene_counts/count_tables/Juvenile_host/all/aten_juveniles_host_all_counts2.csv", row.names = 1))
#view data
#head(x,2)
rownames
rownames(coldata) <- sub("fb","", rownames(coldata))
#Do row names in coldata data match to column data in x
all(rownames(coldata) %in% colnames(x))
#are they in the same order?
all(rownames(coldata) == colnames(x))
#change order
x <- x[, rownames(coldata)]
#double check order
all(rownames(coldata) == colnames(x))
```

```{r}
#Create the matrix
x <- DGEList(counts = x)

# Assuming `sample_info` has a column named `group` for conditions
x$samples$group <- coldata$group

class(x)

dim(x)
```

```{r}
samplenames <- substring(colnames(x), 1, nchar(colnames(x)))
samplenames

colnames(x) <- samplenames
group <- as.factor(c("AmbientD1", "AmbientD1", "AmbientD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "HotD1", "AmbientD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "HotD1", "HotD1"))
x$samples$group <- group
```

```{r}
geneid <- rownames(x)
```

####
Data pre-processing

###
Transformations from the raw-scale
```{r}
cpm <- cpm(x)
lcpm <- cpm(x, log=TRUE)
```

```{r}
L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6
c(L, M)
summary(lcpm)
```

###
Removing genes that are lowly expressed
```{r}
table(rowSums(x$counts==0)==9)

keep.exprs <- filterByExpr(x, group=group)
x <- x[keep.exprs,, keep.lib.sizes=FALSE]
dim(x)
```

```{r}
lcpm.cutoff <- log2(10/M + 2/L)
library(RColorBrewer)
nsamples <- ncol(x)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
lcpm <- cpm(x, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
```

###
Normalising gene expression distributions
```{r}
x <- calcNormFactors(x, method = "TMM")
x$samples$norm.factors
x<- data.matrix(x)
x<- DESeq2::varianceStabilizingTransformation(x)
x<- t(x)
```

```{r}
y<-group
summary(y) # check the distribution of class labels
```

```{r}
x_D1<- x
y_d1<- y
```

##################
Symbionts
Get Count Data
```{r}
x<- as.matrix(read.csv("/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Data/Gene_counts/count_tables/Juvenile_symbiont/aten_juveniles_symb_Dtre_counts.csv", row.names = 1))
#x<- as.matrix(read.csv("/Users/alexmacadam/Dropbox/PhD/Chapter4_RNA/Data/Gene_counts/count_tables/Juvenile_symbiont/aten_#juveniles_symb_Dtre_counts(outlier_filtered).csv", row.names = 1))
#view data
head(x,2)
rownames
rownames(coldata) <- sub("fb","", rownames(coldata))
#Do row names in coldata data match to column data in x
all(rownames(coldata) %in% colnames(x))
#are they in the same order?
all(rownames(coldata) == colnames(x))
#change order
x <- x[, rownames(coldata)]
#double check order
all(rownames(coldata) == colnames(x))
```

```{r}
#Create the matrix
x <- DGEList(counts = x)

# Assuming `sample_info` has a column named `group` for conditions
x$samples$group <- coldata$group
x$samples$Cross <- coldata$Cross

class(x)

dim(x)
```

```{r}
samplenames <- substring(colnames(x), 1, nchar(colnames(x)))
samplenames

colnames(x) <- samplenames
group <- as.factor(c("AmbientD1", "AmbientD1", "AmbientD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "HotD1", "AmbientD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "HotD1", "HotD1"))
#group <- as.factor(c("AmbientD1", "AmbientD1", "AmbientD1", "AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "AmbientD1", #"HotD1", "AmbientD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "AmbientD1", #"HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "HotD1", "AmbientD1", "HotD1", "HotD1", "HotD1", "AmbientD1", #"AmbientD1", "AmbientD1", "HotD1", "AmbientD1", "HotD1"))
x$samples$group <- group

Cross <- as.factor(c("J8xJ2","J8xJ2","J8xD6","J8xJ3","J8xD13","J8xD13","J8xD13","J8xJ3","J8xJ3","J8xD13","J8xD6","J8xJ2","J8xD1","J8xD6","J8xJ2","J8xD1","J8xJ3","J8xD13","J8xD1","J8xD6","J8xJ2","J8xD1","J8xD6","J8xJ3","J8xD1","J8xJ2","J8xD6","J8xJ3","J8xD13","J8xD1","J8xP5","J8xP5","J8xP5","J8xP5","J8xP5","J8xP5"))
x$samples$Cross <- Cross
```

```{r}
geneid <- rownames(x)
```

####
Data pre-processing

###
Transformations from the raw-scale
```{r}
cpm <- cpm(x)
lcpm <- cpm(x, log=TRUE)
```

```{r}
L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6
c(L, M)
summary(lcpm)
```

###
Removing genes that are lowly expressed
```{r}
table(rowSums(x$counts==0)==9)

keep.exprs <- filterByExpr(x, group=group)
x <- x[keep.exprs,, keep.lib.sizes=FALSE]
dim(x)
```

```{r}
lcpm.cutoff <- log2(10/M + 2/L)
library(RColorBrewer)
nsamples <- ncol(x)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
lcpm <- cpm(x, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
```

###
Normalising gene expression distributions
```{r}
x <- calcNormFactors(x, method = "TMM")
x$samples$norm.factors
x<- data.matrix(x)
x<- DESeq2::varianceStabilizingTransformation(x)
x<- t(x)
```

```{r}
y<-group
summary(y) # check the distribution of class labels
```

```{r}
x_Dtre<- x
y_Dtre<- y
```

####################
Data
```{r}
dim(x_D1) # check the dimensions of the X dataframe
dim(x_Dtre) # check the dimensions of the Y dataframe
```

Initial Analysis
Exploration of feature correlation
```{r}
# produce a heat map of the cross correlation matrix
imgCor(x_D1, x_Dtre, sideColors = c("purple", "green"))
```

Initial rCCA model
```{r}
# set grid search values for each regularisation parameter
grid1 <- seq(0.001, 0.2, length = 5) 
grid2 <- seq(0.001, 0.2, length = 5)

# optimise the regularisation parameter values
cv.tune.rcc.nutrimouse <- tune.rcc(x_D1, x_Dtre, grid1 = grid1, grid2 = grid2, 
                                   validation = "loo")
```

```{r}
cv.tune.rcc.nutrimouse # examine the results of CV tuning
```

```{r}
opt.l1 <- cv.tune.rcc.nutrimouse$opt.lambda1 # extract the optimal lambda values
opt.l2 <- cv.tune.rcc.nutrimouse$opt.lambda2

# formed optimised CV rCCA
CV.rcc.nutrimouse <- rcc(X, Y, method = "ridge", 
                         lambda1 = opt.l1, lambda2 = opt.l2) 
```

```{r}
# run the rCCA method using shrinkage
shrink.rcc.nutrimouse <- rcc(X,Y, method = 'shrinkage') 
# examine the optimal lambda values after shrinkage 
shrink.rcc.nutrimouse$lambda
```

```{r}
# barplot of cross validation method rCCA canonical correlations
plot(CV.rcc.nutrimouse, type = "barplot", main = "Cross Validation") 

# barplot of shrinkage method rCCA canonical correlations
plot(shrink.rcc.nutrimouse, type = "barplot", main = "Shrinkage") 
```

Final Model

Sample plots
```{r}
# plot the projection of samples for CV rCCA data
plotIndiv(CV.rcc.nutrimouse, comp = 1:2, 
          ind.names = nutrimouse$genotype,
          group = nutrimouse$diet, rep.space = "XY-variate", 
          legend = TRUE, title = '(a) Nutrimouse, rCCA CV XY-space')

# plot the projection of samples for shrinkage rCCA data
plotIndiv(shrink.rcc.nutrimouse, comp = 1:2, 
          ind.names = nutrimouse$genotype,
          group = nutrimouse$diet, rep.space = "XY-variate", 
          legend = TRUE, title = '(b) Nutrimouse, rCCA shrinkage XY-space')
```

```{r}
# plot the arrow plot of samples for CV rCCA data
plotArrow(CV.rcc.nutrimouse, group = nutrimouse$diet, 
          col.per.group = color.mixo(1:5),
          title = '(a) Nutrimouse, CV method')

# plot the arrow plot of samples for shrinkage rCCA data
plotArrow(shrink.rcc.nutrimouse, group = nutrimouse$diet, 
          col.per.group = color.mixo(1:5),
          title = '(b) Nutrimouse, shrinkage method')
```

Variable plots
```{r}
plotVar(CV.rcc.nutrimouse, var.names = c(TRUE, TRUE),
        cex = c(4, 4), cutoff = 0.5,
        title = '(a) Nutrimouse, rCCA CV comp 1 - 2')

plotVar(shrink.rcc.nutrimouse, var.names = c(TRUE, TRUE),
        cex = c(4, 4), cutoff = 0.5,
        title = '(b) Nutrimouse, rCCA shrinkage comp 1 - 2')
```

```{r}
network(CV.rcc.nutrimouse, comp = 1:2, interactive = FALSE,
        lwd.edge = 2,
        cutoff = 0.5)
```

```{r}
cim(CV.rcc.nutrimouse, comp = 1:2, xlab = "genes", ylab = "lipids")
```

